---
title: "Linear Regression"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{linear_regression}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ucrstats)
```

## Accessing tutorials 

Tutorials can be accessed by placing the tutorial name or "id" into the `access_tutorial` function. The different tutorials are listed below, with the names/ids shown in parentheses. 

```{r, eval=FALSE}
access_tutorial("R_SLR")
```


## Available Tutorials 

*The following are tutorials on linear regression.*

### Simple Linear Regression (R_SLR)

This tutorial provides the basic information on linear relationships between two variables and conducting a simple linear regression model in R.

### Multiple Linear Regression (R_MLR)

This tutorial provides the basic information to conduct a multiple linear regression model in R.

### Linear Regression Residual Analysis (R_Resid)

This tutorial provides the basic code to conduct a residual analysis from a linear regression model.

